---
title: "HOPE_loans_analysis"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#HOPE analysis:

Step 0 : making sure that packages are installed, and environment clean
packages to install in R to be able to run this code:

install.packages("ggplot")
install.packages("caret")
install.packages("glm")
install.packages("readr")
install.packages("tidyr")

obs: I like to load the package further on, as I need to use it: library(), or require()

```{r}
rm(list = ls()) #clear the work environment
```

Step 1: loading the csv into R studio
```{r}
library(readr)
loans <- read_csv("HOPELoansSince2019.csv")
summary(loans)
```

Step 2 : Exploratory analysis
Purpose: to better understand the need for creating variables, and which are the most appropriate models

#Understanding the data
colnames(loans) #Variable names
View(loans) #General overview of the dataset
```{r}
head(loans)
```

#Descriptive statistics of continuous variables: 
#(Min. 1st Qu.  Median    Mean 3rd Qu.    Max.):
```{r}
summary(loans$DisbursedAmount)
summary(loans$LateInstallments)
summary(loans$WorkDaysLoanWasInArrears)
```
#In visual form:
```{r}
boxplot(loans$DisbursedAmount)
boxplot(loans$LateInstallments)
boxplot(loans$WorkDaysLoanWasInArrears)

hist(loans$DisbursedAmount, 500)
hist(loans$LateInstallments, 17)
hist(loans$WorkDaysLoanWasInArrears, 30)

#Inspecting correlation between LateInstallments and WorkDaysLoanWasInASrrears
plot(loans$WorkDaysLoanWasInArrears, loans$LateInstallments)
days_late_model <- lm(loans$LateInstallments ~ loans$WorkDaysLoanWasInArrears )
abline(days_late_model, col = "red")
```
# For categorical/ character variables : use "factor". Now, summary will show  how many in each category
```{r}
summary(factor(loans$Gender))
summary(factor(loans$MaritalStatus))
summary(factor(loans$ProductGroup1Name))
```
# In visual form:

barplot(factor(loans$Gender))
barplot(factor(loans$MaritalStatus))
barplot(factor(loans$ProductGroup1Name))
```{r}
plot(factor(loans$Gender), loans$LateInstallments, main = "Distribution of Late Installements per Gender", xlab = "Gender", ylab = "Number of Late Installements", col="pink")
plot(factor(loans$MaritalStatus), loans$LateInstallments, main = "Distribution of Late Installements per Marital Status", xlab = "Marital Status", ylab = "Number of Late Installements", col="purple")
plot(factor(loans$ProductGroup1Name), loans$LateInstallments, main = "Distribution of Late Installements per Product Group", xlab = "Product Group", ylab = "Number of Late Installements", col="brown")
```
#Dates:
```{r}
summary(factor(loans$BirthDate))
summary(factor(loans$DisbursementDate))
```
#Boxplots with distribution of late installements per month:
plot(factor(loans4$DisbursementDate_YearMonth), loans4$LateInstallments, main = "Distribution of Late Installments per Month", xlab = "Disbursement Date (YearMonth)", ylab = "Number of Late Installments", col="yellow")
```
#just checking, but not too relevant to know its statistics:
summary(loans$CustomerId) # from 100112 to 165421
summary(loans$LoanID)

# Obs: I could have done in one code : summary(loans), but then would need to first create new variables for the categorical ones
#Check: unique LoanID and unique Customer ? No. See distribution of loans per customer
#See histogram
hist(loans$CustomerId, 65421, main="CustomerID", xlab="CustomerID", ylab="Number of loans per customer")

#Count how many customers have a certain number of loans
# insight: create variable to how many loans each customer has had and include that in analysis
# However: be careful since the  individual FE may capture that ? check


#Step 3: Creating new variables###########################

#### AGE:

```{r}
library(tidyr)
loans2 <- separate(data = loans, col = BirthDate, into = c('day_birth', 'month_birth', 'year_birth')) #Separate year from BirthDate column
View(loans2) # checking if columns were correctly created
class(loans2$year_birth) # it is not numeric, but character format
loans2$year <- as.numeric(loans2$year_birth) # transform to numeric
class(loans2$year)#checking : now, it is numeric, ready for calculation
loans2$age <- 2021-loans2$year # create age column
hist(loans2$age) # inspect its distribution
plot(loans2$age, loans2$LateInstallments)
```
#remove outlier? age==89
# PROBLEM: 82% IS MISSING

### DISBURSEMENT DATE: 

```{r}
library(tidyr)
loans3 <- separate(data = loans2, col = DisbursementDate, into = c('month_disbursementDate', 'day_disbursementDate', 'year_disbursementDate')) #Separate year from DisbursementDate column
# Add leading zeros to month
loans3$month_disbursementDate_0 <- paste0("0", loans3$month_disbursementDate)
#Use only the 2 last digits
substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}
loans3$month_Disb.correct <- substrRight(loans3$month_disbursementDate_0, 2)
# use as character still , otherwise, 0 will be dropped! 
#concat just year and month
loans4<-transform(loans3, monthyear_Disbursement.correct=paste0(year_disbursementDate, month_Disb.correct))
loans4$monthyear_Disbursement.num <- as.numeric(loans4$monthyear_Disbursement.correct) # transform to numeric
```
## DELINQUENCY 
```{r}
loans4$Delinquency <- factor(loans4$LateInstallments)
#levels(loans$Delinquency) <- list(None = 0, Low = 1:2, Medium = 3:4, High = 5:18 )
levels(loans4$Delinquency) <- list(Low = 0:2, Medium = 3:4, High = 5:18 )
summary(factor(loans4$Delinquency))
loans4$Delinquency <- ordered(loans4$Delinquency)
```

# Dummies of higher and lower delinquency for probit and logit models:
```{r}
loans4$higher_delinquency <- ifelse(loans4$Delinquency == 'High', 1, 0)
loans4$lower_delinquency <- ifelse(loans4$Delinquency == 'Low', 1, 0)
```
Clean up:
```{r}
loans4$month_disbursementDate_0 <- NULL # dropping
loans4$month_Disb.correct <- NULL
loans4$monthyear_Disbursement.correct <- NULL
loans4$DisbursementDate_YearMonth<-loans4$monthyear_Disbursement.num # renaming
loans4$monthyear_Disbursement.num <- NULL
```
# LOANS PER CUSTOMER:
```{r}
library(dplyr)
loans_per_customer<- tally(group_by(loans4, CustomerId))
loans_per_customer
loans4 <- merge(loans4, loans_per_customer, by = "CustomerId")

```

#Age groups (done):
```{r}
hist(loans4$age)
loans4$age_groups <- factor(loans4$age)
levels(loans4$age_groups) <- list( twenties = 20:29, thirties = 30:39, forties = 40:49, fifties = 50:59, sixties = 60:69)
summary(factor(loans4$age_groups))
loans4$age_groups <- ordered(loans4$age_groups)
summary(loans4$age_groups)
#Remember: NA's because age has NA's
```

####Exploratory analysis after creating variables:

# DISBURSEMENT DATE:
#Exploring the variable:
```{r}
summary(loans4$DisbursementDate_YearMonth)
summary(factor(loans4$DisbursementDate_YearMonth))
plot(factor(loans4$DisbursementDate_YearMonth), loans4$LateInstallments)
plot(factor(loans4$DisbursementDate_YearMonth))
```
#inspect outliers for each variable:
#Boxplot

```{r}
library(ggplot2)
ggplot(loans,aes(x=ProductGroup1Name,y=LateInstallments,
                 fill=factor(ProductGroup1Name)))+
  geom_boxplot()+
  labs(title="Boxplot for Late Installments per ProductGroup1Name",x="ProductGroup1Name",y="Late Installments")+
  theme_bw()+coord_flip()+
  theme(legend.position = "none")

ggplot(loans,aes(x=MaritalStatus,y=LateInstallments,
                 fill=factor(MaritalStatus)))+
  geom_boxplot()+
  labs(title="Boxplot for Late Installments per MaritalStatus",x="MaritalStatus",y="Late Installments")+
  theme_bw()+coord_flip()+
  theme(legend.position = "none")

ggplot(loans,aes(x=Gender,y=LateInstallments,
                 fill=factor(Gender)))+
  geom_boxplot()+
  labs(title="Boxplot for Late Installments per Gender",x="Gender",y="Late Installments")+
  theme_bw()+coord_flip()+
  theme(legend.position = "none")

ggplot(loans4,aes(x=age_groups,y=LateInstallments,
                 fill=factor(age_groups)))+
  geom_boxplot()+
  labs(title="Boxplot for Late Installments per Age Group",x="Age Group",y="Late Installments")+
  theme_bw()+coord_flip()+
  theme(legend.position = "none")

ggplot(loans4,aes(x=age_groups,y=LateInstallments,
                  fill=factor(age_groups)))+
  geom_boxplot()+
  labs(title="Boxplot for Late Installments per Age Group",x="Age Group",y="Late Installments")+
  theme_bw()+coord_flip()+
  theme(legend.position = "none")
```
Save changes in a csv: 
```{r}
write.csv(loans4,"HOPE_dataset_loans4_updated.csv" )
```

# Make date variable : day, month, year #inspect how it plots

#Additional variables that would be relevant if available:
#occupation
#the use of the loan (ex: agriculture, shop, sewing business, etc)
#number of people in household
#location (country, city, )

#################### FIXING variable PROBLEMS:
##PROBLEM: MISSING BIRTHDATE for a huge portion of the dataset
```{r}
summary(factor(loans$BirthDate)) # 82% IS NA
```
# how to deal?
OPTION 1: JUST USE NON-MISSING in age FOR THIS ANALYSIS
```{r}
withage<- loans4[!is.na(loans4$age),] # subset for non missing on age
summary(withage)
write.csv(subset,"HOPE_dataset_withage.csv" )
```
# Ideally: inspect if the occurrence of missing values is related with some characteristics
# In other words: can this subset be considered a random sample of the population ?
#(obs: inspect for other columns?)

# OPTION 2: CHECK HOW GOOD THE EXPLANATORY VARIABLES ARE IN PREDICTING, TO SEE IF IMPUTATION WOULD WORK
# See how good is the subset for prediction

#split using k-fold, dividing into 2:
```{r}
require(caret)
flds <- createFolds( subset$Delinquency, k = 2, list = TRUE, returnTrain = FALSE)
names(flds)[1] <- "train"
training <- subset[flds$train,]
testing <- subset[ flds[[2]], ]
```
### Decision Tree

```{r}
install.packages("e1071")
library(e1071)
ModFit_rpart <- train(Delinquency~.,data=training,method="rpart",
                      parms = list(split = "gini"))
predict_rpart <- predict(ModFit_rpart,testing)
confusionMatrix(predict_rpart, testing$Delinquency)
```
# If adequate, then:
# method 1: set NA to mean
```{r}
NA2mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
new_loans4 <-replace(loans4, TRUE, lapply(loans4, NA2mean))
summary(factor(new_loans4$age))
```
# method 2: imputation with Bootstrap Aggregation Imputation:
```{r}
library(caret)
PreImputeBag <- preProcess(loans4,method="bagImpute")
DataImputeBag <- predict(PreImputeBag,loans4)
```
# method 3: imputation via knn
```{r}
install.packages("RANN")
library(RANN)
MData <- loans4[,-c(1,5,6)]
PreImputeKNN <- preProcess(MData,method="knnImpute",k=5)
DataImputeKNN <- predict(PreImputeKNN,MData)
#Convert back to original scale
RescaleDataM <- t(t(DataImputeKNN)*PreImputeKNN$std+PreImputeKNN$mean)
```

######################################################## extra:


# Disbursement_amounts_categories: PROBLEM: SHOULDN'T HAVE NA'S
#Inspecting again to see what would be reasonable:
```{r}
hist(loans4$DisbursedAmount, 200)
boxplot(loans4$DisbursedAmount)
summary(loans4$DisbursedAmount)
sum(is.na(loans4$DisbursedAmount)) # There are no missing values on DisbursedAmount
#Decision: use the quantiles
# Creating:
loans4$amount_levels <- factor(loans4$DisbursedAmount)
levels(loans4$amount_levels) <- list( first_q = 0:200000, second_q = 200000:311060, third_q = 311060:350000, forth_q = 350000:6000000)
summary(factor(loans4$amount_levels))
loans4$amount_levels <- ordered(loans4$amount_levels)
summary(loans4$amount_levels)
```
#Age at disbursement (to explore):
```{r}
loans2$date.new <- as.Date(as.character(loans$BirthDate), format="%m/%d/%Y")
loans2$date.now <- as.Date(as.character(loans$DisbursementDate), format="%m/%d/%Y")
loans2$age_at_disbusement <- loans$date.new - loans$date.now
loans2$age <- as.numeric(loans2$DisbursementDate - loans2$BirthDate) %/% 365.25
```   

